Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2024-02-20 paper/main.tex
checking in updates to merge from upstream

    % TODO: adjust width to max possible based on finalized caption
    \includegraphics[width=0.75\textwidth]{figs/predict-knowledge-questions}
    \caption{\textbf{Predicting success on held-out questions using estimated
    knowledge.} We used generalized linear mixed models (GLMMs) to model the
    likelihood of correctly answering a quiz question as a function of
    estimated knowledge for its embedding coordinate (see
    \nameref{subsec:glmm}). Separately for each quiz (column), we examined this
    relationship based on three different sets of knowledge estimates:
    knowledge for each question based on all other questions the same
    participant answered on the same quiz (``All questions''; top row),
    knowledge for each question about one lecture based on all questions (from
    the same participant and quiz) about the \textit{other} lecture
    (``Across-lecture''; middle rows), and knowledge for each question about
    one lecture based on all other questions (from the same participant and
    quiz) about the \textit{same} lecture (``Within-lecture''; bottom rows).
    The background in each panel displays the kernel density estimates of the
    relative observed proportions of correctly (blue) versus incorrectly (red)
    answered questions, for each level of estimated knowledge along the
    $x$-axis. The black curves display the (population-level) GLMM-predicted
    probabilities of correctly answering a question as a function of estimated
    knowledge. Error ribbons denote bootstrap-estimated 95\% confidence
    intervals. \textbf{JRM NOTE: RE-ORDER ROWS TO MATCH THE TEXT, AND UPDATE
    CAPTION ACCORDINGLY}}

@@

In performing these analyses, our null hypothesis is that the knowledge
estimates we compute based on the quiz questions' embedding coordinates do
\textit{not} provide useful information about participants' abilities to answer
those questions. What result might we expect to see if this is the case? To
gain an intuition for this possibility, consider the expected outcome if we
carried out these same analyses using a simple proportion-correct measure in
lieu of our knowledge estimates. Suppose a participant correctly answered $n$
out of $q$ questions on a given quiz. If we hold out a single
\textit{correctly} answered question, the proportion of remaining questions
answered correctly would be $\frac{n - 1}{q - 1}$. If we hold out a single
\textit{incorrectly} answered question, the proportion of remaining questions
answered correctly would be $\frac{n}{q - 1}$. In this way, the proportion of
correctly answered remaining questions is always \textit{lower} when the
held-out question was answered correctly than when it was answered incorrectly.
Because our knowledge estimates are computed as a weighted version of this same
proportion-correct score (where each held-in question's weight reflects its
embedding-space distance from the held-out question; see Eqn.~\ref{eqn:prop}),
if these weights were uninformative (e.g., randomly distributed), then we
should expect to see this same inverse relationship between estimated knowledge
and performance, on average. On the other hand, if the spatial relationships
among the quiz questions' embeddings \textit{are} predictive of participants'
knowledge about the questions' content, then we would expect \textit{higher}
estimated knowledge for held-out correctly (versus incorrectly) answered
questions.

When we fit a GLMM to estimates of participants' knowledge for each Quiz~1
question based on all other Quiz~1 questions, we observed an outcome consistent
with our null hypothesis: higher estimated knowledge at the embedding
coordinate of a held-out question was associated with a lower likelihood of
answering the question correctly (odds ratio $(OR) = 0.136$, likelihood-ratio
test statistic $(\lambda_{LR}) = 19.749$, 95\%\ $\textnormal{CI} = [14.352,\
26.545],\ p = 0.001$). This outcome suggests that our knowledge estimates do
\textit{not} provide useful information about participants' Quiz~1 performance
when we aggregated across all question content areas. We speculated that this
might either indicate that the knowledge estimates are uninformative in
general, or about Quiz~1 performance in particular. When we repeated this
analysis for Quizzes~2 and~3, the direction of this relationship reversed:
higher estimated knowledge for a given question predicted a greater likelihood
of answering it correctly (Quiz~2: $OR = 2.905,\ \lambda_{LR} = 17.333,\ 95\%\
\textnormal{CI} = [14.966,\ 29.309],\ p = 0.002$; Quiz~3: $OR = 3.238,\
\lambda_{LR} = 6.882,\ 95\%\ \textnormal{CI} = [6.228,\ 8.184],\ p = 0.017$).
Taken together, these results suggest that our knowledge estimates \textit{do}
reliably predict participants' likelihood of success on individual quiz
questions, but only when there is some ``contrast'' in performance between the
content of different questions. On Quiz~1, before participants have learned the
content of either lecture, their overall performance is poor (average
proportion correct: $0.46$), and the specific questions they answer correctly
not not seem to be structured in a meaningful way with respect to the embedding
space (e.g., Fig.~\ref{fig:knowledge-maps}A, left panel). On Quizzes~2 and~3,
after participants have begun their training, their overall performance
improves (average proportion correct for Quiz~2: 0.60; average proportion
correct for Quiz~3: 0.79), and the specific content areas they improve on are
captured by the lecture and question embeddings (and, therefore, our knowledge
estimates).

@@

In the set of analyses reported in Figure~\ref{fig:predictions}, we applied Equation~\ref{eqn:prop} to

estimate participants' knowledge for individual held-out quiz questions (i.e., treating

In the set of analyses reported in Figure~\ref{fig:predictions}, we applied Equation~\ref{eqn:prop} to estimate participants' knowledge at the embedding coordinates of individual held-out quiz questions. For each of the three quizzes a participant took,

Equation~\ref{eqn:prop} to estimate participants' knowledge at the embedding coordinate of individual held-out questions, based on other questions

all other questions (``All questions''; Fig.~\ref{fig:predictions}, top row),

(``All questions''; Fig.~\ref{fig:predictions}, top row).
This test was intended to assess the overall predictive power of our approach. 
Second, we estimated knowledge for each question about a given lecture using only the other questions (from the same participant and quiz) about that \textit{same} lecture (``Within-lecture''; Fig.~\ref{fig:predictions}, middle rows). 
This test was intended to assess the \textit{specificity} of our approach by asking whether our predictions could distinguish between questions about different content covered by the same lecture.
Third, we estimated knowledge for each question about one lecture using only questions (from the same participant and quiz) about the \textit{other} lecture (``Across-lecture''; Fig.~\ref{fig:predictions}, bottom rows).

To analyze the predictive value of

To predict individual participants' success on held-out quiz questions, we used Equation~\ref{eqn:prop}

@@

%\begin{lstlisting}[language=R]
%    accuracy \sim knowledge + (knowledge|participant) + (knowledge|question)
%\end{lstlisting}

@@

>>>>

    % TODO: adjust width to max possible based on finalized caption
    \includegraphics[width=0.75\textwidth]{figs/predict-knowledge-questions}
    \caption{\textbf{Predicting success on held-out questions using estimated
    knowledge.} We used generalized linear mixed models (GLMMs) to model the
    likelihood of correctly answering a quiz question as a function of
    estimated knowledge for its embedding coordinate (see
    \nameref{subsec:glmm}). Separately for each quiz (column), we examined this
    relationship based on three different sets of knowledge estimates:
    knowledge for each question based on all other questions the same
    participant answered on the same quiz (``All questions''; top row),
    knowledge for each question about one lecture based on all other questions 
    (from the same participant and quiz) about the \textit{same} lecture 
    (``Within-lecture''; middle rows), and knowledge for each question about 
    one lecture based on all questions (from the same participant and quiz) 
    about the \textit{other} lecture (``Across-lecture''; bottom rows).
    The background in each panel displays the kernel density estimates of the
    relative observed proportions of correctly (blue) versus incorrectly (red)
    answered questions, for each level of estimated knowledge along the
    $x$-axis. The black curves display the (population-level) GLMM-predicted
    probabilities of correctly answering a question as a function of estimated
    knowledge. Error ribbons denote the 95\% confidence
    intervals.}

@@

In performing these analyses, our null hypothesis is that the knowledge
estimates we compute based on the quiz questions' embedding coordinates do
\textit{not} provide useful information about participants' abilities to answer
those questions. What result might we expect to see if this is the case? To
gain an intuition for this possibility, consider the expected outcome if we
carried out these same analyses using a simple proportion-correct measure in
lieu of our knowledge estimates. Suppose a participant correctly answered $n$
out of $q$ questions on a given quiz. If we hold out a single
\textit{correctly} answered question, the proportion of remaining questions
answered correctly would be $\frac{n - 1}{q - 1}$. Whereas if we hold out a single
\textit{incorrectly} answered question, the proportion of remaining questions
answered correctly would be $\frac{n}{q - 1}$. In this way, the proportion of
correctly answered remaining questions is always \textit{lower} when the
held-out question was answered correctly than when it was answered incorrectly.
Because our knowledge estimates are computed as a weighted version of this same
proportion-correct score (where each held-in question's weight reflects its
embedding-space distance from the held-out question; see Eqn.~\ref{eqn:prop}),
if these weights are uninformative (e.g., randomly distributed), then we
should expect to see this same inverse relationship between estimated knowledge
and performance, on average. On the other hand, if the spatial relationships
among the quiz questions' embeddings \textit{are} predictive of participants'
knowledge about the questions' content, then we would expect \textit{higher}
estimated knowledge for held-out correctly (versus incorrectly) answered
questions.

When we fit a GLMM to estimates of participants' knowledge for each Quiz~1
question based on all other Quiz~1 questions, we observed an outcome consistent
with our null hypothesis: higher estimated knowledge at the embedding
coordinate of a held-out question was associated with a lower likelihood of
answering the question correctly (odds ratio $(OR) = 0.136$, likelihood-ratio
test statistic $(\lambda_{LR}) = 19.749$, 95\%\ $\textnormal{CI} = [14.352,\
26.545],\ p = 0.001$). This outcome suggests that our knowledge estimates do
\textit{not} provide useful information about participants' Quiz~1 performance
when we aggregated across all question content areas. We speculated that this
might either indicate that the knowledge estimates are uninformative in
general, or about Quiz~1 performance in particular. When we repeated this
analysis for Quizzes~2 and~3, the direction of this relationship reversed:
higher estimated knowledge for a given question predicted a greater likelihood
of answering it correctly (Quiz~2: $OR = 2.905,\ \lambda_{LR} = 17.333,\ 95\%\
\textnormal{CI} = [14.966,\ 29.309],\ p = 0.002$; Quiz~3: $OR = 3.238,\
\lambda_{LR} = 6.882,\ 95\%\ \textnormal{CI} = [6.228,\ 8.184],\ p = 0.017$).
Taken together, these results suggest that our knowledge estimates \textit{do}
reliably predict participants' likelihood of success on individual quiz
questions, but only when there is some ``contrast'' in performance between the
content of different questions. On Quiz~1, before participants have learned the
content of either lecture, their overall performance is poor (average
proportion correct: $0.46$), and the specific questions they answer correctly
do not seem to be structured in a meaningful way with respect to the embedding
space (e.g., Fig.~\ref{fig:knowledge-maps}A, left panel). On Quizzes~2 and~3,
after participants have begun their training, their overall performance
improves (average proportion correct for Quiz~2: 0.60; average proportion
correct for Quiz~3: 0.79), and the specific content areas they improve on are
captured by the lecture and question embeddings (and, therefore, our knowledge
estimates).

@@

In the set of analyses reported in Figure~\ref{fig:predictions}, we assessed whether estimates of participants' knowledge at the embedding coordinates of individual quiz questions could reliably predict their ability to correctly answer those questions. 
In essence, we treated each question a given participant answered on a given quiz as a ``lecture'' consisting of a single timepoint, and used Equation~\ref{eqn:prop} to estimate the participant's knowledge for its embedding coordinate based on their performance on all \textit{other} questions they answered on that same quiz (``All questions''; Fig.~\ref{fig:predictions}, top row). 
We also computed analogous knowledge estimates for each lecture-related question based on all other questions (answered by the same participant, on the same quiz) about the same lecture (``Within-lecture''; Fig.~\ref{fig:predictions}, middle rows) and based on all questions about the other of the two lectures (``Across-lecture''; Fig.~\ref{fig:predictions}, bottom rows).

In each version of this analysis (i.e., row in Fig.~\ref{fig:predictions}), and separately for each of the three quizzes (i.e., column in Fig.~\ref{fig:predictions}),  we then fit a generalized linear mixed model (GLMM) with a logistic link function to the set of knowledge estimates for all questions answered by all participants on the given quiz.
Our models were implemented in \texttt{R} using the \texttt{lme4} package \citep{BateEtal15}.

%In each versions of this analysis (i.e., row in Fig.~\ref{fig:predictions}), and separately for each of the three quizzes (i.e., column in Fig.~\ref{fig:predictions}),  we then fit a generalized linear mixed model (GLMM) with a logistic link function to the set of

%In the set of analyses reported in Figure~\ref{fig:predictions}, we used generalized linear mixed models (GLMMs) to analyze whether estimates of participants' knowledge at the embedding coordinates of individual held-out questions could reliably predict their ability to answer those questions correctly.

@@

@@

$$
\begin{aligned}
  \operatorname{accuracy}_{i}  &\sim \operatorname{Binomial}(n = 1, \operatorname{prob}_{\operatorname{accuracy} = \operatorname{correct}} = \widehat{P}) \\
    \log\left[\frac{\hat{P}}{1 - \hat{P}} \right] &=\alpha_{j[i],k[i]} \\    
\left(
  \begin{array}{c} 
    \begin{aligned}
      &\alpha_{j} \\
      &\gamma_{1j}
    \end{aligned}
  \end{array}
\right)
  &\sim N \left(
\left(
  \begin{array}{c} 
    \begin{aligned}
      &\gamma_{0}^{\alpha} + \gamma_{1k[i]}^{\alpha}(\operatorname{knowledge}) \\
      &\mu_{\gamma_{1j}}
    \end{aligned}
  \end{array}
\right)
, 
\left(
  \begin{array}{cc}
     \sigma^2_{\alpha_{j}} & \rho_{\alpha_{j}\gamma_{1j}} \\ 
     \rho_{\gamma_{1j}\alpha_{j}} & \sigma^2_{\gamma_{1j}}
  \end{array}
\right)
 \right)
    \text{, for participant j = 1,} \dots \text{,J} \\    
\left(
  \begin{array}{c} 
    \begin{aligned}
      &\alpha_{k} \\
      &\gamma_{1k}
    \end{aligned}
  \end{array}
\right)
  &\sim N \left(
\left(
  \begin{array}{c} 
    \begin{aligned}
      &\mu_{\alpha_{k}} \\
      &\mu_{\gamma_{1k}}
    \end{aligned}
  \end{array}
\right)
, 
\left(
  \begin{array}{cc}
     \sigma^2_{\alpha_{k}} & \rho_{\alpha_{k}\gamma_{1k}} \\ 
     \rho_{\gamma_{1k}\alpha_{k}} & \sigma^2_{\gamma_{1k}}
  \end{array}
\right)
 \right)
    \text{, for question k = 1,} \dots \text{,K}
\end{aligned}
$$
